Learning Multiple Networks via Supervised Tensor Decomposition

### 2020-10-02 tensor_regress/NeurIPSw/neuripsw_tensor_v2.tex
update

\begin{abstract}
   We develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike classical tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating optimization algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project. We identify the key global brain connectivity pattern and pinpoint the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas.
\end{abstract}

\section{Introduction}
 Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. A popular example is in neuroimaging~\citep{zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in network analysis~\citep{pmlr-v108-berthet20a,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent  friendships. Side information such as people’s demographic information and friendship types are often available. In both examples, it is of keen scientific interest to identify the variation in the data tensor (e.g., brain connectivities, social community patterns) that is affected by available features. These seemingly different scenarios pose a common yet challenging problem for tensor data modeling.

 \begin{figure}[hbt]
\begin{center}
\includegraphics[width = 13.5cm]{demo.pdf}
\vspace{-.5cm}
\end{center}
\caption{Examples of supervised tensor decomposition with interactive side information. (a) Network population model. (b) Spatio-temporal growth model.} \label{fig:intro1}
\vspace{-.3cm}
\end{figure}

@@

We present a general model and associated method for decomposing a data tensor whose entries are from exponential family with interactive side information. We formulate the learning task as a structured regression problem, with tensor observation serving as the response, and the multiple side information as interactive features. We leverage generalized linear model (GLM)  to allow heteroscedacity due to the mean-variance relationship in the non-Gaussian data. The low-rank structure on the conditional mean tensor effectively mitigates the curse of high dimensionality.  Our proposal blends the modeling power of GLM and the exploratory capability of tensor dimension reduction in order to take the best out of both worlds. 
%A low-rank structure is imposed to the conditional mean of tensor observation.where unlike classical decomposition, the tensor factors $\mX_k\mM_k\in\mathbb{R}^{d_k\times r_k}$ belong to the space spanned by features $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,2,3$. The unknown matrices $\mM_k\in\mathbb{R}^{p_k\times r_k}$ (referred to as ``dimension reduction matrices'') link the conditional mean to the feature spaces, thereby allowing the identification of variations in the tensor data attributable to the side information.

@@

\vspace{-0.3cm}

\section{Method}

\subsection{Preliminary}
We introduce the basic tensor properties used in the paper. We use lower-case letters (e.g.,\ $a,b,c$) for scalars and vectors, upper-case boldface letters (e.g.,\ $\mA, \mB, \mC$) for matrices, and calligraphy letters (e.g.,\ $\tA, \tB, \tC$) for tensors of order three or greater. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as $\tY\times\{\mX_1,\ldots,\mX_K\} =\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\cdots x_{j_K,i_K}^{(K)}}$,
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor.For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as $\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}.$
The tensor Frobenius norm is defined as $\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}$, and the maximum norm is defined $\mnormSize{}{\tY}=\max_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}.$  We let $\mI_d$ denote the $d \times d$ identity matrix and $[d]$ denote the $d$-set $\{1,\ldots,d\}$.
%For any two matrices $\mA,\mB$ with orthonormal columns of same dimension, the angle distance is defined as
%\[ \sin \Theta(\mA,\mB)=\snormSize{}{\mA^T \mB^{\perp}}=\max\left\{ \frac{\langle \mx, \my\rangle}{\vnormSize{}{\mx}\vnormSize{}{\my}}\colon \ \mx\in \textup{Span}(\mA),\ \my\in \textup{Span}(\mB^\perp)\right\}. \]

\subsection{General Model}

Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. Assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent realizations from an exponential family distribution, and the conditional mean tensor admits the form
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\quad \text{with}\quad \Theta =\tB\times\{\mX_1,\ldots, \mX_K\} ,
\end{align}
where $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the unknown parameter tensor, $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$. The choice of link function is based on the assumed distribution family of tensor entries.

In classical tensor decomposition, tensor factorization is performed on either data tensor $\tY$ or mean tensor $\mathbb{E}(\tY)$. In the context of supervised tensor decomposition, we propose to factorize the latent parameter tensor $\tB$,
\begin{equation}\label{eq:factor}
\tB=\tC\times\{\mM_1,\ \ldots,\ \mM_K\},
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$. By the definition of multilinear rank, model equations~\eqref{eq:tensormodel} and~\eqref{eq:factor} imply the low-rankness $\mr=(r_1,\ldots,r_K)$ of the conditional mean tensor under the link function. We now reach our final model for supervised tensor decomposition,
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1, \ldots, \mX_K\mM_K\}\right),\notag\\
\text{with} \ &\ \mM_k^T\mM_k = \mI_{r_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\quad \text{for all }k=1,\ldots,K,
\end{align}
where the parameters of interest are $\mM_k$ and $\tC$. Note that model~\eqref{eq:decomp} assumes a fixed, known rank $\mr=(r_1,\ldots,r_K)$. Figure~\ref{fig:intro1}b provides a schematic illustration of our model. The features $\mX_k$ affect the distribution of tensor entries in $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. We call $\mX_k\mM_k$ the ``supervised tensor factors'' or ``sufficient features''~\citep{adragni2009sufficient}, and call $\mM_k$ the ``dimension reduction matrix.'' The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes, which links the conditional mean to the feature spaces, and thereby allows the identification of variations in the tensor data attributable to the side information. Our goal is to find $\mM_k$ and the corresponding $\tC$,thereby allowing us to reveal the relationship between side information $\mX$ and the observed tensor $\tY$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

\vspace{-.3cm}
\section{Estimation}
\subsection{Rank-constrained M-estimator}
 We adopt the exponential family as a flexible framework for different data types. In a classical generalized linear model with a scalar response $y$ and feature $\mx$, the density is expressed as
\[
p(y|\mx, \boldsymbol{\beta})=c(y,\phi)\exp\left(\frac{y\theta- b(\theta)}{\phi}\right)\ \text{with}\ \theta=\boldsymbol{\beta}^T\mx,
\]
where $b(\cdot)$ is a known function depending on the data types, $\theta$ is the linear predictor, $\phi>0$ is the dispersion parameter, and $c(\cdot)$ is a known normalizing function. In our context, we model the tensor entries $y_{i_1,\ldots,i_K}$, conditional on $\theta_{i_1,\ldots,i_K}$, as independent draws from an exponential family. Ignoring constants that do not depend on $\Theta$, the quasi log-likelihood of~\eqref{eq:decomp} is equal to Bregman distance between $\tY$ and $b'(\Theta)$:
\begin{equation}\label{eq:loglikelihood}
\small
\tL_{\tY}(\tC,\mM_1,\ldots,\mM_K)=\langle \tY, \Theta \rangle - \sum\nolimits_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K})\ \text{with}\ \Theta=\tC\times\{\mM_1\mX_1,\ldots,\mM_K\mX_K\}.
\end{equation}
We propose a constrained maximum quasi-likelihood estimator (M-estimator),
\begin{align} \label{eq:MLE} 
(\hat \tC, \hat \mM_1,\ldots,\hat \mM_k) &=\argmax_{(\tC,\mM_1,\ldots,\mM_K)\in \tP} \ \tL_{\tY}(\tC,\mM_1,\ldots,\mM_K),
\end{align}
where parameter space 
$
\tP=\left\{(\tC, \mM_1,\ldots,\mM_K) \ \Big| \ \mM_k^T\mM_k=\mathbf{I}_{r_k},\ \mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha \right\}.
$
The maximum norm constraint on the linear predictor $\Theta$ is a technical condition to avoid the divergence in the non-Gaussian variance.

\subsection{Alternating optimization}

@@

\subsection{Statistical properties}\label{subsec:statprob}

In modern applications, the tensor data and features are often large-scale. We are particularly interested in the high-dimensional regime in which both $d_k$ and $p_k$ diverge; i.e.\ $d_k\to \infty$ and $p_k\to\infty$, while $p_k/d_k \to \gamma_k \in[0,1)$. As the size of problem grows, and so does the number of unknown parameters. The classical MLE theory does not directly apply. We leverage the recent development in random tensor theory and high-dimensional statistics to establish the error bounds.

\begin{thm}[Statistical convergence]\label{thm:main}
Consider a data tensor generated from model~\eqref{eq:decomp}. Let $(\hat \tC, \hat \mM_1,\ldots,\hat \mM_K)$ be the M-estimator in~\eqref{eq:MLE} and $\hat \tB=\hat \tC\times \hat \mM_1\times\cdots \times \hat \mM_K$. Define $r_\textup{total}=\prod_k r_k$ and $r_{\max}=\max_k r_k$. Under mild technical assumptions, there exist two positive constants $C_1=C_1(\alpha,K), C_2=C_2(\alpha, K)>0$ independent of dimensions $\{d_k\}$ and $\{p_k\}$, such that, with probability at least $1-\exp(-C_1\sum_k p_k)$, 
\begin{equation}\label{eq:bound}
    \FnormSize{}{\trueB- \hat \tB}^2\leq \frac{ C_{2} r_{\textup{total}}}{r_{\max}}\frac{\sum_k p_k}{\prod_k d_k}.
\end{equation}
Furthermore, if the unfolded core tensor has non-degenerate singular values at mode $k\in[K]$, i.e., $\sigma_{\min}(\textup{Unfold}_k(\tC_\textup{true})) \geq c>0$ for some constant $c$, then
\begin{equation}\label{eq:sinebound}
\textup{sin}^2 \Theta(\trueM,\ \hat \mM_k) \leq  \frac{ C_{2}r_{\textup{total}}}{ r_{\max}\sigma^2_{\min}(\textup{Unfold}_k(\tC_\textup{true}))}\frac{\sum_k p_k}{\prod_k d_k},
\end{equation}
where $
\sin \Theta(\trueM,\hat \mM_k)=\snormSize{}{\trueM^T \hat \mM_k^{\perp}}=\max\left\{ \frac{\langle \mx, \my\rangle}{\vnormSize{}{\mx}\vnormSize{}{\my}}\colon \ \mx\in \textup{Span}(\trueM),\ \my\in \textup{Span}({\hat \mM_k}^\perp)\right\}
$ is the angle distance to assess the accuracy in estimating the column space, $\textup{Span}(\mM_k)$.
\end{thm}

\section{Real data analysis}

The Human Connectome Project (HCP) aims to build a network map that characterizes the anatomical and functional connectivity within healthy human brains~\citep{HCP}. 
%We analyze the structural connectivity patterns among 68 brain regions for 136 individuals from HCP. 
We follow the preprocessing procedure as in~\cite{zhang2018mapping} and parcellate the brain into 68 regions of interest~\citep{desikan2006automated}. The dataset consists of 136 brain structural networks, one for each individual. Each brain network is represented as a 68-by-68 binary matrix, where the entries encode the presence or absence of fiber connections between the 68 brain regions. We consider four individual features: gender (65 females vs.\ 71 males), age 22-25 ($n=35$), age 26-30 ($n=58$), and age 31+ ($n=43$). The goal is to identify the connection edges that are affected by individual features. 
%A key challenge in brain network is that the edges are correlated; for example, the nodes in edges may be from a same brain region, and it is of importance to take into account the within-dyad dependence.

\begin{figure}[!h]
\centering
\includegraphics[width=14cm]{HCP.pdf}
\caption{Top edges with large effects. (a) Global effect; (b) Female effect; (c) Age 22-25; (d) Age 31+. Red edges represent positive effects and blue edges represent negative effects. The edge-width is proportional to the magnitude of the effect size.
}\label{fig:brain}
%\vspace{1cm}
%\includegraphics[width=14cm]{compare_HCP.pdf}
%\caption{Comparison of estimated feature effects the HCP data using (a) multi-response GLM and (b) supervised tensor decomposition (STD). }\label{fig:s1}
%\vspace{-.1cm}
\end{figure}

We perform the supervised tensor decomposition to the HCP data. %The data tensor is binary, $\tY\in\{0,1\}^{68\times 68\times 136}$, and the features are of dimension 4 on the 3$^{\text{rd}}$ mode. 
The BIC selection suggests a rank $\mr=(10,10,4)$ with quasi log-likelihood $\tL_{\tY}=-174654.7$. We utilize the sum-to-zero contrasts in coding the feature effects, and depict only the top 3\% edges whose connections are non-constant across the sample. Figure~\ref{fig:brain} shows the top edges with high effect size, overlaid on the Desikan atlas brain template~\citep{desikan2006automated}. We find that the global connection exhibits clear spatial separation, and that the nodes within each hemisphere are more densely connected with each other (Figure~\ref{fig:brain}a). In particular, the superior-temproal (\emph{SupT}), middle-temporal (\emph{MT}) and Insula are the top three popular nodes in the network. Interestingly, female brains display higher inter-hemispheric connectivity, especially in the frontal, parental and temporal lobes (Figure~\ref{fig:brain}b). This is in agreement with a recent study showing that female brains are optimized for inter-hemispheric communication~\citep{ingalhalikar2014sex}. We find several edges with declined connection in the group Age 31+. Those edges involve Frontal-pole (\emph{Fploe}), superior-frontal (\emph{SupF}) and Cuneus nodes. The Frontal-pole region is known for its importance in memory and cognition, and the detected decline with age further highlights its biological importance.

@@

Our supervised tensor decomposition method is widely applicable to network analysis, dyadic data analysis, spatial-temporal model, and recommendation systems. We have shown the improved predictive power and enhanced interpretability by incorporating the interactive side information in tensor decomposition method.  The application to the brain connection dataset yields conclusions with sensible interpretations, suggesting the practical utility of the proposed approach.  We believe that our model enriches the research of tensor-based learning and is a powerful tool to boost scientific discoveries in various fields.

>>>>

\begin{abstract}
We consider the problem of tensor decomposition with multiple side information available as interactive features. Such problems are common in neuroimaging, network modeling, and spatial-temporal analysis. We develop a new family of exponential tensor decomposition models and establish the theoretical accuracy guarantees. An efficient alternating optimization algorithm is further developed. Unlike earlier methods, our proposal handles a broad range of data types, including continuous, count, and binary observations, along with available features. We apply the method to diffusion tensor imaging data from human connectome project and identify the key brain connectivity patterns associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, all data and code has been made available to the public. 
\end{abstract}

\section{Introduction}
 Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. A popular example is in neuroimaging~\citep{zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in network analysis~\citep{pmlr-v108-berthet20a}. Side information such as people’s demographic information and friendship types are often available. In both examples, scientists are interested in identifing the variation in the data tensor (e.g., brain connectivities, social community patterns) that is affected by available features. These seemingly different scenarios pose a common yet challenging problem for tensor data modeling.

 \begin{figure}[hbt]
 \vspace{-.4cm}
\begin{center}
\includegraphics[width = 13.5cm]{demo.pdf}
\vspace{-.4cm}
\end{center}
\caption{Examples of supervised tensor decomposition with interactive side information. (a) Network population model. (b) Spatio-temporal growth model.} \label{fig:intro1}
\vspace{-.2cm}
\end{figure}

@@

{\bf Our contribution.} This paper presents a general model and associated method for decomposing a data tensor whose entries are from exponential family with interactive side information. We formulate the learning task as a low-rank tensor regression problem, with tensor observation serving as the response, and the multiple side information as interactive features. We blend the modeling power of generalized linear model (GLM) and the exploratory capability of tensor dimension reduction in order to take the best out of both worlds. Our methods greatly improves the classical tensor decomposition, and we quantify the gain in prediction through numerical experiments and data applications.

{\bf Notation.} We follow the tensor notation as in~\cite{kolda2009tensor}. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as $\tY\times\{\mX_1,\ldots,\mX_K\} =\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\cdots x_{j_K,i_K}^{(K)}}$,
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. The inner product between two tensors of equal size is defined as $\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}$. For ease of notation, we allow basic arithmetic operators (e.g.,\ $+, -$) and univariate functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner.

%A low-rank structure is imposed to the conditional mean of tensor observation.where unlike classical decomposition, the tensor factors $\mX_k\mM_k\in\mathbb{R}^{d_k\times r_k}$ belong to the space spanned by features $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,2,3$. The unknown matrices $\mM_k\in\mathbb{R}^{p_k\times r_k}$ (referred to as ``dimension reduction matrices'') link the conditional mean to the feature spaces, thereby allowing the identification of variations in the tensor data attributable to the side information.

@@

\section{Proposed models and motivating examples}
%For any two matrices $\mA,\mB$ with orthonormal columns of same dimension, the angle distance is defined as
%\[ \sin \Theta(\mA,\mB)=\snormSize{}{\mA^T \mB^{\perp}}=\max\left\{ \frac{\langle \mx, \my\rangle}{\vnormSize{}{\mx}\vnormSize{}{\my}}\colon \ \mx\in \textup{Span}(\mA),\ \my\in \textup{Span}(\mB^\perp)\right\}. \]
Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. We assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent realizations from an exponential family distribution, and the conditional mean tensor admits the form
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1, \ldots, \mX_K\mM_K\}\right),\notag\\
\text{with} \ &\ \mM_k^T\mM_k = \mI_{r_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\quad \text{for all }k=1,\ldots,K.
\end{align}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$, $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$. Common choices of link functions include identity link for Gaussian distribution, logistic link for Bernoulli distribution, and $\exp(\cdot)$ link for Poisson distribution.

Figure~\ref{fig:intro1}b provides a schematic illustration of our model. The features $\mX_k$ affect the distribution of tensor entries in $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes, which links the conditional mean to the feature spaces, and thereby allows the identification of variations in the tensor data attributable to the side information. Our goal is to find $\mM_k$ and the corresponding $\tC$, thereby allowing us to reveal the relationship between side information $\mX_k$ and the observed tensor $\tY$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

We give two examples of supervised tensor decomposition models~\eqref{eq:decomp} that arise in practice.
\begin{example}[Spatio-temporal growth model]
The growth curve model~\citep{srivastava2008models} was originally proposed as an example of bilinear model for matrix data, and we extend it to higher-order cases. Let $\tY=\entry{y_{ijk}}\in\mathbb{R}^{d \times m\times n}$ denote the pH measurements of $d$ lakes at $m$ levels of depth and for $n$ time points. Suppose the sampled lakes belong to $q$ types, with $p$ lakes in each type. Let $\{\ell_j\}_{j\in[m]}$ denote the sampled depth levels and $\{t_k\}_{k\in[n]}$ the time points. Assume that the expected pH trend in depth is a polynomial of order at most $r$ and that the expected trend in time is a polynomial of order $s$. Then, the conditional mean model for the spatio-temporal growth is a special case of our model~\eqref{eq:decomp}, where $\mX_1=\text{blockdiag}\{\mathbf{1}_q,\ldots,\mathbf{1}_q\}\in \{0,1\}^{d\times p}$ is the design matrix for lake types, 
\[
\mX_2=
\begin{pmatrix}
1 & \ell_1&\cdots &\ell^{r}_1\\
1 & \ell_2&\cdots &\ell^{r}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&\ell_{m}&\cdots&\ell^{r}_{m}
\end{pmatrix},\
\mX_3=
\begin{pmatrix}
1 & t_1&\cdots &t^{s}_1\\
1 & t_2&\cdots &t^{s}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&t_{n}&\cdots&t^{s}_{n}
\end{pmatrix}
\]
are the design matrices for spatial and temporal effects, respectively. The spatial-temporal mode has covariates available on each of the three modes.

\begin{example}[Network population model] 
Network response model is recently developed in the context of neuroimanig analysis. The goal is to study the relationship between network-valued response and the individual covariates. Suppose we observe $n$ i.i.d.\ observations $\{(\mY_i, \mx_i): i=1,\ldots,n\}$, where $\mY_i\in\{0,1\}^{d\times d}$ is the brain connectivity network on the $i$-th individual, and $\mx_i\in\mathbb{R}^p$ is the individual covariate such as age, gender, cognition, etc. The network-response model~\citep{rabusseau2016low} has the form
\begin{equation}\label{eq:network}
\text{logit}(\mathbb{E}(\mY_i|\mx_i))=\tB\times_3\mx_i, \quad \text{for }i=1,\ldots,n
\end{equation}
where $\tB\in \mathbb{R}^{d\times d\times p}$ is the coefficient tensor of interest. The model~\eqref{eq:network} is also a special case of our tensor-response model, with covariates on the last mode of the tensor.

\section{Estimation algorithms}
We develop a likelihood-based procedure to estimate $\tC$ and $\mM_k$ in~\eqref{eq:decomp}. Ignoring constants that do not depend on $\Theta$, the quasi log-likelihood of~\eqref{eq:decomp} is equal to
\begin{equation}\label{eq:loglikelihood}
\small
\tL_{\tY}(\tC,\mM_1,\ldots,\mM_K)=\langle \tY, \Theta \rangle - \sum\nolimits_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K})\ \text{with}\ \Theta=\tC\times\{\mM_1\mX_1,\ldots,\mM_K\mX_K\},
\end{equation}
where $b(\theta)=\theta^2/2$ for Gaussian response, $b(\theta)=\exp(\theta)$ for Poisson response, and $b(\theta)=\log(1+\exp(\theta))$ for Bernoulli response. We propose a constrained maximum quasi-likelihood estimator (M-estimator),
\begin{align} \label{eq:MLE} 
(\hat \tC, \hat \mM_1,\ldots,\hat \mM_k) &=\argmax_{(\tC,\mM_1,\ldots,\mM_K)\in \tP} \ \tL_{\tY}(\tC,\mM_1,\ldots,\mM_K),
\end{align}
where parameter space 
$
\tP=\left\{(\tC, \mM_1,\ldots,\mM_K) \ \Big| \ \mM_k^T\mM_k=\mathbf{I}_{r_k},\ \mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha \right\}.
$

@@

We provide the accuracy guarantee for the proposed M-estimator~\eqref{eq:MLE} by leveraging recent development in random tensor theory and high-dimensional statistics.
\begin{thm}[Convergence]\label{thm:main}
Let $(\hat \tC, \hat \mM_1,\ldots,\hat \mM_K)$ be the M-estimator in~\eqref{eq:MLE} and $\hat \tB=\hat \tC\times \hat \mM_1\times\cdots \times \hat \mM_K$. Define $r_\textup{total}=\prod_k r_k$ and $r_{\max}=\max_k r_k$. Under mild technical assumptions, there exist two positive constants $C_1, C_2>0$, such that, with probability at least $1-\exp(-C_1\sum_k p_k)$, 
\begin{equation}\label{eq:bound}
    \FnormSize{}{\trueB- \hat \tB}^2\leq \frac{ C_{2} r_{\textup{total}}}{r_{\max}}\frac{\sum_k p_k}{\prod_k d_k},\quad \text{and}\quad
\sin^2 \Theta(\trueM,\ \hat \mM_k) \leq  \frac{ C_{2}r_{\textup{total}}}{ r_{\max}\sigma^2_{\min}(\textup{Unfold}_k(\tC_\textup{true}))}\frac{\sum_k p_k}{\prod_k d_k},
\end{equation}
where $\sin \Theta(\trueM,\hat \mM_k)=\snormSize{}{\trueM^T \hat \mM_k^{\perp}}$ is the angle distance between column spaces. 
\end{thm}
Theorem~\ref{thm:main} implies that the estimation has a convergence rate $\tO(d^{-(K-1)})$ in the special case when tensor dimensions are equal on each of the modes, i.e., $d_k=d$ for all $k\in[K]$, and feature dimension grows with tensor dimension, $p_k=\gamma d$, $\gamma\in [0,1)$, for $k\in[K]$. The convergence of our estimation becomes especially favorable as the order of tensor data increases.

\section{Numerical experiments}
We compare our supervised tensor decomposition ({\bf STD}) with three other supervised tensor methods: Higher-order low-rank regression ({\bf HOLRR}~\cite{rabusseau2016low}), Higher-order partial least square ({\bf HOPLS}~\cite{zhao2012higher}) and Subsampled tensor projected gradient ({\bf TPG}~\cite{yu2016learning}). Figure~\ref{fig:compare} shows that {\bf STD} outperforms others, especially in the low-signal, high-rank setting. As the number of informative modes (i.e., modes with available features) increases, the {\bf STD} exhibits a substantial reduction in error whereas others remain unchanged (Figure~\ref{fig:compare}b). This showcases the benefit of incorporation of multiple features. The accuracy gain in Figure~\ref{fig:compare} demonstrates the benefit of alternating algorithm -- incorporation of informative modes also improves the estimation in the non-informative modes.

\begin{figure}[ht]
\centering
\includegraphics[width=14cm]{compare_alternative_w.pdf} 
\vspace{-.5cm}
\caption{\small Comparison between different tensor methods. Panels (a) and (b) plot mean squared prediction error (MSPE) versus the number of modes with available features. Panels (c) and (d) plot MSPE versus the effective sample size $d^2$. We consider rank $\mr=(3,3,3)$ (low) vs (4,5,6)
 (high), and signal $\alpha =3 $ (low) vs.\ 6 (high).}~\label{fig:compare}
\vspace{-.5cm}
\end{figure}

We then apply our method to brain structural connectivity networks from Human Connectome Project (HCP)~\citep{HCP}. %We analyze the structural connectivity patterns among 68 brain regions for 136 individuals from HCP. 
The dataset consists of 136 brain structural networks, one for each individual. Each brain network is represented as a 68-by-68 binary matrix, where the entries encode the presence or absence of fiber connections between the 68 brain regions. We consider four individual features: gender (65 females vs.\ 71 males), age 22-25 ($n=35$), age 26-30 ($n=58$), and age 31+ ($n=43$). The goal is to identify the connection edges that are affected by individual features. 
%A key challenge in brain network is that the edges are correlated; for example, the nodes in edges may be from a same brain region, and it is of importance to take into account the within-dyad dependence.

\begin{figure}[!h]
\centering
\includegraphics[width=14cm]{HCP.pdf}
\vspace{-.5cm}
\caption{\small Top edges with large effects. (a) Global effect; (b) Female effect; (c) Age 22-25; (d) Age 31+. Red (blue) edges represent positive (negative) effects. Edge-widths are proportional to the magnitudes of effect sizes.
}\label{fig:brain}
\vspace{-.2cm}
%\includegraphics[width=14cm]{compare_HCP.pdf}
%\caption{Comparison of estimated feature effects the HCP data using (a) multi-response GLM and (b) supervised tensor decomposition (STD). }\label{fig:s1}
%\vspace{-.1cm}
\end{figure}

We perform the supervised tensor decomposition to the HCP data. The BIC selection suggests a rank $\mr=(10,10,4)$ with quasi log-likelihood $\tL_{\tY}=-174654.7$. Figure~\ref{fig:brain} shows the top edges with high effect size, overlaid on the Desikan atlas brain template~\citep{desikan2006automated}. We find that the global connection exhibits clear spatial separation, and that the nodes within each hemisphere are more densely connected with each other (Figure~\ref{fig:brain}a). In particular, the superior-temproal (\emph{SupT}), middle-temporal (\emph{MT}) and Insula are the top three popular nodes in the network. Interestingly, female brains display higher inter-hemispheric connectivity, especially in the frontal, parental and temporal lobes (Figure~\ref{fig:brain}b). This is in agreement with a recent study showing that female brains are optimized for inter-hemispheric communication~\citep{ingalhalikar2014sex}. We find several edges with declined connection in the group Age 31+. Those edges involve Frontal-pole (\emph{Fploe}), superior-frontal (\emph{SupF}) and Cuneus nodes. Our results highlight the importance of Frontal-pole region, and the detected decline further suggests the age effects to brain connections.

\section{Conclusion}
\vspace{-.2cm}
We have developed a supervised tensor decomposition method with side information on multiple modes. The empirical results demonstrate the improved interpretability and accuracy over previous approaches. Applications to the brain connection data yield conclusions with sensible interpretations, suggesting the practical utility of the proposed approach.

@@

Our supervised tensor decomposition method is widely applicable to network analysis, dyadic data analysis, spatial-temporal model, and recommendation systems. We have shown the improved predictive power and enhanced interpretability by incorporating the interactive side information in tensor decomposition method.  The application to the brain connection dataset yields conclusions with sensible interpretations, suggesting the practical utility of the proposed approach.  Tensor learning is a clear challenge for further research. We believe that our model enriches the research of tensor-based learning and is a powerful tool to boost scientific discoveries in various fields. We hope the work opens up new inquiry that allows more machine learning researchers to contribute to this field.
